\section{Reproducibility}
\label{app:reproducibility}

We provide full details necessary to reproduce every experiment
reported in this paper.

\subsection{Code Availability}

All source code is publicly available at
\url{https://github.com/gaius050526/torch-eml} under the MIT license.
The library, \texttt{torch-eml} v0.1.0, is implemented entirely in
PyTorch and can be installed directly from the repository.

\subsection{Library Structure}

The codebase is organized into the following key modules:

\begin{itemize}
  \item \texttt{torch\_eml.primitives} --- Verified EML constructions
    providing the atomic building blocks of the framework:
    \texttt{EMLExp}, \texttt{EMLLn}, \texttt{EMLSin},
    \texttt{EMLCos}, and \texttt{EMLPi}.
  \item \texttt{torch\_eml.compose} --- Higher-level architectures for
    composing primitives into multi-variable expressions, including
    \texttt{ComposeHead}, \texttt{SeparableTerm}, and
    \texttt{AxisTerm}.
  \item \texttt{torch\_eml.symbolic} --- Symbolic extraction and
    snapping utilities that convert trained network parameters into
    closed-form SymPy expressions.
\end{itemize}

\subsection{Dependencies}

\begin{itemize}
  \item PyTorch $\geq$ 2.0
  \item SymPy $\geq$ 1.12
\end{itemize}

No additional dependencies beyond standard Python scientific libraries
are required.

\subsection{Hardware}

All experiments reported in this paper were run on a single CPU (Apple
M-series). No GPU is required. The complete Taylor--Green vortex search,
including both the search and fine-tuning phases, completes in
approximately two minutes on this hardware.

\subsection{Random Seed}

All experiments use a fixed random seed for reproducibility:
\begin{verbatim}
  torch.manual_seed(42)
\end{verbatim}

\subsection{Hyperparameters for the Taylor--Green Experiment}

Table~\ref{tab:hyperparams} summarizes the hyperparameters used for
the Taylor--Green vortex experiment.

\begin{table}[h]
\centering
\caption{Hyperparameters for the Taylor--Green vortex experiment.}
\label{tab:hyperparams}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Training data points     & 3000 \\
Spatial domain           & $x, y \sim \mathcal{U}[0,\, 2\pi]$ \\
Temporal domain          & $t \sim \mathcal{U}[0,\, 1]$ \\
Function basis           & $\{\sin,\, \cos,\, \exp\}$ \\
\midrule
\multicolumn{2}{l}{\textit{Search phase}} \\
\quad Steps              & 2000 per candidate \\
\quad Optimizer          & Adam \\
\quad Learning rate      & 0.01 \\
\quad Gradient clipping  & 1.0 \\
\midrule
\multicolumn{2}{l}{\textit{Fine-tune phase}} \\
\quad Steps              & 5000 \\
\quad Optimizer          & Adam \\
\quad Learning rate      & 0.005 \\
\quad Gradient clipping  & 1.0 \\
\midrule
Snap tolerance           & 0.05 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Test Suite}

The repository includes a comprehensive test suite comprising 117
tests that cover all modules, including primitive correctness,
composition semantics, symbolic extraction, and end-to-end integration.
All tests pass under the released version of the library.
