\section{Experiments}
\label{sec:experiments}

We evaluate the proposed pipeline on seven classes of problems: separable PDE
benchmarks spanning one- and two-dimensional domains
(Section~\ref{sec:separable_benchmarks}), a non-separable traveling-wave
recovery that demonstrates multi-term search
(Section~\ref{sec:nonseparable}), robustness and failure-mode analysis
(Section~\ref{sec:robustness}), ablation studies isolating each design
choice (Section~\ref{sec:ablation}), PDE-residual-only discovery that
recovers exact solutions from the governing equations alone without any
solution data (Section~\ref{sec:pde_residual}), a systematic
three-dimensional Navier--Stokes screening that tests which classes of 3-D
initial conditions admit exact separable solutions
(Section~\ref{sec:3d_ns}), and recovery of nonlinear PDE solutions using
an extended hyperbolic basis (Section~\ref{sec:hyperbolic}).
Section~\ref{sec:summary_table} collects
all results in a unified comparison table.  All experiments use
PyTorch~2.0+~\citep{paszke2019} on a single CPU; the most expensive search
(81 two-term pairs for the traveling wave) completes in under ten minutes.

% ===========================================================================
\subsection{Separable PDE Benchmarks}
\label{sec:separable_benchmarks}

We begin with three single-term separable PDEs whose exact solutions are
products of axis-aligned elementary functions.  In each case the pipeline
recovers the ground truth to within machine precision after snapping.

% ---------------------------------------------------------------------------
\subsubsection{1-D Heat Equation}
\label{sec:heat}

The one-dimensional heat equation $u_t = \nu\,u_{xx}$ on $x\in[0,1]$,
$t\in[0,1]$ with $\nu = 1/\pi^{2}$ admits the exact solution
\begin{equation}
  u(x,t) = \sin(\pi x)\,\exp(-\pi^{2}\nu\, t)
         = \sin(\pi x)\,\exp(-t).
  \label{eq:heat_exact}
\end{equation}

\paragraph{Setup.}
We sample $n=2{,}000$ training points uniformly over the domain and set the
function basis to $\mathcal{F}=\{\sin,\cos,\exp\}$, yielding
$|\mathcal{F}|^2 = 9$ two-factor separable candidates of the form
$f(x)\,g(t)$.  Each candidate is trained for 2{,}000 steps with Adam
(learning rate $0.01$, gradient clipping at max-norm~$1.0$); the best
candidate is then fine-tuned for 5{,}000 steps at learning rate $0.005$.

\paragraph{Result.}
The search selects $\sin(\cdot)\!\cdot\!\exp(\cdot)$ as the lowest-loss
candidate (loss $< 10^{-9}$ after fine-tuning).  After normalization and
snapping, the recovered expression is
$\hat{u}(x,t) = \sin(\pi x)\,\exp(-t)$,
with snap MAE of $9\times 10^{-8}$ on 5{,}000 held-out points.

% ---------------------------------------------------------------------------
\subsubsection{1-D Wave Equation}
\label{sec:wave}

The one-dimensional wave equation $u_{tt} = c^{2}\,u_{xx}$ on $x\in[0,1]$,
$t\in[0,2]$ with $c=1$ and initial conditions $u(x,0)=\sin(\pi x)$,
$u_t(x,0)=0$ has the standing-wave solution
\begin{equation}
  u(x,t) = \sin(\pi x)\,\cos(\pi t).
  \label{eq:wave_exact}
\end{equation}

\paragraph{Setup.}
The configuration is identical to the heat equation: $n=2{,}000$ training
points, basis $\mathcal{F}=\{\sin,\cos,\exp\}$, 9 candidates, 2{,}000 search
steps followed by 5{,}000 fine-tuning steps.

\paragraph{Result.}
The winning candidate is $\sin(\cdot)\!\cdot\!\cos(\cdot)$ with loss
$< 10^{-9}$.  Snapping yields
$\hat{u}(x,t)=\sin(\pi x)\,\cos(\pi t)$
with snap MAE of $5\times 10^{-8}$.

% ---------------------------------------------------------------------------
\subsubsection{2-D Navier--Stokes: Taylor--Green Vortex}
\label{sec:tgv}

The Taylor--Green vortex is an exact, time-decaying solution to the
two-dimensional incompressible Navier--Stokes equations~\citep{taylor1937}:
\begin{align}
  u(x,y,t) &= \phantom{-}\sin(x)\cos(y)\,\exp(-2\nu t), \label{eq:tgv_u}\\
  v(x,y,t) &= -\cos(x)\sin(y)\,\exp(-2\nu t), \label{eq:tgv_v}\\
  p(x,y,t) &= -\tfrac{1}{4}\bigl[\cos(2x)+\cos(2y)\bigr]\exp(-4\nu t),
              \label{eq:tgv_p}
\end{align}
where $\nu=0.1$ is the kinematic viscosity.  Because the velocity field is a
product of separable, axis-aligned elementary functions, this problem provides
an ideal test of whether the \textsc{ComposeHead} can recover exact symbolic
structure from data alone.

\paragraph{Setup.}
We sample $n=3{,}000$ training points uniformly over
$x,y\in[0,2\pi]$, $t\in[0,1]$ and evaluate both velocity components at the
exact solution~\eqref{eq:tgv_u}--\eqref{eq:tgv_v}.  The function basis is
$\mathcal{F}=\{\sin,\cos,\exp\}$, yielding $|\mathcal{F}|^3=27$ triple
separable candidates of the form $f(x)\,g(y)\,h(t)$.  Each candidate is
trained independently for 2{,}000 steps with Adam (learning rate $0.01$,
gradient clipping at max-norm~$1.0$).  The best candidate for each velocity
component is then fine-tuned for an additional 5{,}000 steps at learning rate
$0.005$.

\paragraph{Search results.}
For the $u$-component, the search identifies
$\sin(\cdot)\!\cdot\!\sin(\cdot)\!\cdot\!\exp(\cdot)$ as the lowest-loss
candidate (loss $<10^{-8}$); for $v$, the winner is
$\cos(\cdot)\!\cdot\!\cos(\cdot)\!\cdot\!\exp(\cdot)$.  These selections appear
to differ from the ground truth, which contains $\sin\!\cdot\!\cos$ and
$\cos\!\cdot\!\sin$ factors, respectively.  The discrepancy is resolved by
observing that the optimizer absorbs phase shifts into the bias parameters:
$\sin(x - \pi)\sin(y - \pi/2)\exp(at+b) = \sin(x)\cos(y)\exp(at)\exp(b)$
up to sign, so the learned representation is equivalent to the true solution
modulo a reparameterization that the normalization step removes.

\paragraph{Raw learned parameters.}
After fine-tuning, the combined MSE loss reaches $\sim\!10^{-10}$.  The raw
parameters before normalization are as follows:
\begin{itemize}
  \item $u$: $\text{coeff}=0.540$;\;
        $\sin_0$: $a\!=\!1.000$, $b\!=\!{-3.142}\;(\approx -\pi)$;\;
        $\sin_1$: $a\!=\!1.000$, $b\!=\!{-1.571}\;(\approx -\pi/2)$;\;
        $\exp_2$: $a\!=\!{-0.200}$, $b\!=\!0.616$.
  \item $v$: $\text{coeff}=0.594$;\;
        $\cos_0$: $a\!=\!1.000$, $b\!=\!0.000$;\;
        $\cos_1$: $a\!=\!1.000$, $b\!=\!1.571\;(\approx \pi/2)$;\;
        $\exp_2$: $a\!=\!{-0.200}$, $b\!=\!0.522$.
\end{itemize}

\paragraph{Normalization and snapping.}
The normalization step (i)~absorbs exponential biases into the coefficient
($e^{0.616}\approx 1.851$, yielding $0.540 \times 1.851 \approx 1.000$),
(ii)~canonicalizes trigonometric phases by applying the identities
$\sin(x-\pi)=-\sin(x)$ and $\sin(y-\pi/2)=\cos(y-\pi)\mapsto -\cos(y)$
together with the sign-absorption rule for $\sin$, and
(iii)~reduces all bias terms to zero.
After normalization, coefficient snapping with tolerance~$0.05$ yields the
exact closed-form expressions:
\begin{equation}
  \hat{u} = \sin(x)\cos(y)\exp(-t/5), \qquad
  \hat{v} = -\cos(x)\sin(y)\exp(-t/5).
  \label{eq:recovered}
\end{equation}
The snap error---defined as the mean absolute difference between the snapped
model and the ground truth on 5{,}000 held-out points---is $7\times10^{-8}$,
consistent with machine precision.

\paragraph{PDE residual verification.}
We verify the recovered solution against the governing equations.  The
continuity residual $|\nabla\!\cdot\!\mathbf{u}|$ has mean value
$3\times10^{-4}$, attributable to finite-difference approximation of the
divergence.  The momentum residuals are non-zero because the pipeline does not
model the pressure field; the residual in each momentum equation equals
$\nabla p$, which is consistent with the pressure
gradient implied by~\eqref{eq:tgv_p}.

\paragraph{Comparison with baselines.}
Table~\ref{tab:tgv_comparison} compares four approaches on the same
Taylor--Green data set.

\begin{table}[t]
  \centering
  \caption{Comparison of methods on the Taylor--Green vortex.  ``Data Error''
    is the mean squared error on the training data after convergence.
    ``Symbolic Output'' indicates whether the method produces a human-readable
    formula.}
  \label{tab:tgv_comparison}
  \smallskip
  \begin{tabular}{@{}lccc@{}}
    \toprule
    Method & Data Error & Symbolic Output & Interpretable \\
    \midrule
    MLP-PINN (3-layer, 64-wide)
      & ${\sim}\,10^{-3}$ & No (weight matrix) & No \\
    Raw EML tree (depth 4)
      & ${\sim}\,0.28$    & Opaque nested \texttt{eml()} & Barely \\
    \textsc{ComposeHead} (free weights)
      & ${\sim}\,10^{-3}$ & Messy multi-term sum & Partially \\
    \textsc{ComposeHead} (separable)
      & $7\!\times\!10^{-8}$ & $\sin(x)\cos(y)\exp(-t/5)$ & \textbf{Yes} \\
    \bottomrule
  \end{tabular}
\end{table}

The MLP-PINN~\citep{raissi2019} achieves low data error but yields an opaque
weight matrix.  The raw EML tree (depth~4) struggles with the three-variable
product structure and converges to a poor local minimum.  The
\textsc{ComposeHead} with free (non-axis-aligned) weight vectors fits the data
reasonably but distributes signal across multiple terms, producing a multi-term
expression that resists simplification.  Only the separable
\textsc{ComposeHead} recovers the exact symbolic solution.

% ===========================================================================
\subsection{Non-Separable Recovery via Multi-Term Search}
\label{sec:nonseparable}

The single-term separable ansatz $f(x)\,g(t)$ cannot represent functions whose
natural decomposition requires a sum of separable products.  A canonical
example is the traveling wave
\begin{equation}
  u(x,t) = \sin(x - 2t),
  \label{eq:traveling}
\end{equation}
which by the angle-subtraction identity decomposes as
\begin{equation}
  u(x,t) = \sin(x)\cos(2t) - \cos(x)\sin(2t).
  \label{eq:traveling_decomp}
\end{equation}
This two-term structure lies strictly outside the single-product function class
but within the class of sums of separable products.  Recovering it tests
whether the pipeline can extend beyond the single-term setting.

\paragraph{Setup.}
We sample $n=2{,}000$ training points uniformly over $x\in[0,2\pi]$,
$t\in[0,2]$.  The function basis is $\mathcal{F}=\{\sin,\cos,\exp\}$, giving
$|\mathcal{F}|^2=9$ possible single-term candidates per factor.  For
the two-term search, we enumerate all $9\times 9 = 81$ ordered pairs of
function assignments, where each pair consists of two separable products
$c_1\,f_1(x)\,g_1(t) + c_2\,f_2(x)\,g_2(t)$.  Each pair is trained jointly
for 2{,}000 steps with Adam (learning rate $0.01$), and the best pair is
fine-tuned for 5{,}000 additional steps at learning rate $0.005$.

\paragraph{Bias freezing.}
A critical implementation detail is that \emph{all bias parameters are frozen
at zero} during the multi-term search.  The reason is a gauge freedom inherent
to multi-term decompositions: if the two terms sum to a fixed target, then
correlated phase offsets---such as adding $\delta$ to the bias of $f_1$ and
simultaneously adjusting $c_1$, $c_2$, and the bias of $f_2$ to
compensate---leave the total sum unchanged while preventing any individual
term from snapping to a clean symbolic form.  In the single-term setting this
gauge freedom is absent because the normalization step can absorb all phase
offsets into the coefficient and canonicalize each factor independently.
Freezing biases at zero eliminates the multi-term gauge manifold and forces
the optimizer to express the target using canonical basis functions directly.

\paragraph{Result.}
The single-term search correctly identifies that no single product achieves low
loss: all 9 candidates plateau at loss $> 0.13$ (see
Section~\ref{sec:robustness}).  In the two-term search, the winning pair is
$\sin(x)\cos(t) + \cos(x)\sin(t)$, achieving loss $< 10^{-8}$ after
fine-tuning.  After coefficient snapping, the recovered expression is
\begin{equation}
  \hat{u}(x,t) = \sin(x)\cos(2t) - \cos(x)\sin(2t),
  \label{eq:traveling_recovered}
\end{equation}
with snap MAE of $3\times 10^{-8}$ on 5{,}000 held-out points---again
consistent with machine precision.  This confirms that the separable ansatz
extends naturally to \emph{sums} of separable terms, covering a strictly larger
function class than single-product solutions.

% ===========================================================================
\subsection{Robustness and Failure Analysis}
\label{sec:robustness}

A trustworthy symbolic-recovery method should not only succeed on clean data
but also degrade gracefully under adverse conditions.  We examine three
scenarios: observation noise, a misspecified function basis, and a
single-term search applied to a non-separable target.

% ---------------------------------------------------------------------------
\subsubsection{Noise Robustness}
\label{sec:noise}

We add i.i.d.\ Gaussian noise $\varepsilon \sim \mathcal{N}(0,\sigma^2)$ to
the Taylor--Green training data and repeat the full pipeline (27-candidate
search, fine-tuning, normalization, snapping) for each noise level.
Table~\ref{tab:noise} reports the snap MAE of the recovered $u$-component.

\begin{table}[t]
  \centering
  \caption{Noise robustness on the Taylor--Green $u$-component.  For
    $\sigma \le 0.1$ the method recovers the exact symbolic expression; at
    $\sigma = 0.2$ recovery fails.}
  \label{tab:noise}
  \smallskip
  \begin{tabular}{@{}lcc@{}}
    \toprule
    Noise level $\sigma$ & Snap MAE & Exact recovery? \\
    \midrule
    $0.0$   & $7\!\times\!10^{-8}$ & Yes \\
    $0.001$ & $8\!\times\!10^{-8}$ & Yes \\
    $0.01$  & $6\!\times\!10^{-8}$ & Yes \\
    $0.05$  & $4\!\times\!10^{-7}$ & Yes \\
    $0.1$   & $9\!\times\!10^{-7}$ & Yes \\
    $0.2$   & ---                   & No  \\
    \bottomrule
  \end{tabular}
\end{table}

At noise levels $\sigma \le 0.1$ (corresponding to a signal-to-noise ratio of
roughly 10 or greater, given that the Taylor--Green velocity field has unit
amplitude), the correct candidate is selected and the snapped parameters match
the ground truth.  The snap MAE increases modestly with $\sigma$ but remains
below $10^{-6}$---well within the snapping tolerance.  At $\sigma = 0.2$ the
noise overwhelms the signal, the search selects an incorrect candidate, and
snapping fails.  The method thus tolerates moderate observation noise, a
property it inherits from the continuous relaxation: the gradient-based
optimizer is a natural denoiser that averages over the training set during
each update.

% ---------------------------------------------------------------------------
\subsubsection{Wrong Basis}
\label{sec:wrong_basis}

We repeat the Taylor--Green search with a deliberately misspecified basis
$\mathcal{F}' = \{\exp, \ln\}$, omitting all trigonometric functions.  This
yields $|\mathcal{F}'|^3 = 8$ candidates.  Every candidate plateaus at
training loss $> 0.18$, far above the $< 10^{-8}$ threshold achieved by the
correct basis.  Importantly, the method does not produce a spurious
``recovery'': the large residual loss signals unambiguously that no candidate
in the pool is adequate.  This behavior is desirable---a symbolic regression
system should fail loudly when the hypothesis class does not contain the
target, rather than returning a misleading approximate fit
\citep{cranmer2023}.

% ---------------------------------------------------------------------------
\subsubsection{Non-Separable Target with Single-Term Search}
\label{sec:nonsep_failure}

Applying the single-term separable search to the traveling wave
$u(x,t) = \sin(x - 2t)$ provides a complementary failure mode.  All 9
candidates of the form $f(x)\,g(t)$ converge to training losses above $0.13$,
correctly indicating that no single separable product can represent the
target.  This gap between the single-term plateau ($>0.13$) and the two-term
optimum ($<10^{-8}$) provides a reliable diagnostic: when the best single-term
loss is orders of magnitude above machine precision, the user is directed to
attempt a multi-term search (Section~\ref{sec:nonseparable}).

% ===========================================================================
\subsection{Ablation Study}
\label{sec:ablation}

We isolate the effect of each design decision on the Taylor--Green benchmark.
Table~\ref{tab:ablation} summarizes the results.

\begin{table}[t]
  \centering
  \caption{Ablation study on the Taylor--Green vortex.  Each row modifies one
    component of the default pipeline (separable single-term search with
    normalization).  ``Snap Error'' is the mean absolute error of the snapped
    model on 5{,}000 held-out points.}
  \label{tab:ablation}
  \smallskip
  \begin{tabular}{@{}lll@{}}
    \toprule
    Variant & Snap Error & Outcome \\
    \midrule
    \textbf{Full pipeline (default)}
      & $7\!\times\!10^{-8}$ & Exact recovery \\
    Free-weight terms (non-separable)
      & ${\sim}\,10^{-3}$ & Multi-term local minimum \\
    Multi-term with $L_1$ regularization
      & ${\sim}\,10^{-2}$ & Over-pruning of 60+ competing terms \\
    Without normalization (direct snap)
      & $0.005$--$0.02$ & Phase offsets corrupt snapping \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Separable vs.\ free-weight terms.}
When each factor's weight vector is a free $\mathbb{R}^3$ vector rather than
an axis-aligned scalar, the optimizer converges to a representation that
spreads signal across multiple terms.  The resulting multi-term sum achieves
reasonable data error (${\sim}\,10^{-3}$) but does not simplify to the
known single-product expression, because the weight sharing across input
dimensions introduces a degenerate manifold of equivalent parameterizations.

\paragraph{Single-term search vs.\ multi-term with $L_1$.}
A natural alternative to our candidate-by-candidate search is to instantiate
all 27 separable terms simultaneously and apply $L_1$ regularization to the
coefficients to encourage sparsity~\citep{brunton2016}.  In practice, this
approach over-prunes: with 60+ competing terms (27 for each velocity component
plus cross-terms), the $L_1$ penalty suppresses the correct term before the
optimizer can distinguish it from its neighbors.  Single-term search avoids
this failure mode by giving each candidate the full optimization budget in
isolation.

\paragraph{Effect of normalization.}
Without the normalization step, snapping must operate directly on the raw
learned parameters, which carry arbitrary phase offsets
($b \approx -\pi$, $-\pi/2$) and exponential biases ($b \approx 0.6$).
Rounding these values to the nearest ``clean'' constants introduces errors of
$0.005$--$0.02$, two to five orders of magnitude worse than snapping after
normalization.  The normalization step is therefore essential for exact
recovery.

\paragraph{Candidate pool coverage.}
Among the 27 candidates searched for each velocity component, the top~4
candidates all achieve final loss below $10^{-7}$.  These correspond to the
four function triples that are related to the ground truth by the
$\sin\leftrightarrow\cos$ phase-shift equivalence (e.g.,
$\sin\!\cdot\!\sin\!\cdot\!\exp$ and $\cos\!\cdot\!\cos\!\cdot\!\exp$ for the
$u$-component both encode $\sin(x)\cos(y)\exp(-t/5)$ under suitable phase
offsets).  The remaining 23 candidates converge to losses above $10^{-2}$,
providing a clear separation that makes candidate selection unambiguous.

% ===========================================================================
\subsection{PDE-Residual-Only Discovery}
\label{sec:pde_residual}

All preceding experiments supply solution data---values of the target function
at sampled points---and use standard regression loss to drive training.  A
more ambitious question is whether the pipeline can discover exact analytical
solutions from the governing equations alone, without \emph{any} solution
data.  This is the setting of physics-informed neural networks
(PINNs)~\citep{raissi2019}, but with a critical difference: the output is a
closed-form symbolic expression rather than a set of neural-network weights.

\paragraph{Formulation.}
We consider the two-dimensional incompressible Navier--Stokes equations in
vorticity form, which eliminates the pressure variable:
\begin{equation}
  \omega_t + u\,\omega_x + v\,\omega_y = \nu\,(\omega_{xx} + \omega_{yy}),
  \qquad
  u_x + v_y = 0,
  \qquad
  \omega = v_x - u_y,
  \label{eq:vorticity_ns}
\end{equation}
where $\nu = 0.1$.  The loss function contains no solution data; it consists
entirely of PDE residuals evaluated at randomly sampled collocation points
plus an initial-condition penalty:
\begin{equation}
  \mathcal{L} = \mathcal{L}_{\mathrm{PDE}} + \lambda\,\mathcal{L}_{\mathrm{IC}},
  \label{eq:pde_loss}
\end{equation}
where $\mathcal{L}_{\mathrm{PDE}}$ is the mean-squared vorticity-equation and
continuity residuals computed via automatic differentiation, and
$\mathcal{L}_{\mathrm{IC}}$ penalizes deviation from the Taylor--Green
initial conditions $u_0 = \sin(x)\cos(y)$, $v_0 = -\cos(x)\sin(y)$ at
$t = 0$.

\paragraph{Setup.}
The search space consists of 27 separable candidates for $u$ (basis
$\mathcal{F} = \{\sin,\cos,\exp\}$, three factors for $x$, $y$, $t$).  For
each $u$-candidate, the $v$-candidate is constructed by swapping $\sin
\leftrightarrow \cos$ in the $x$ and $y$ factors, motivated by the continuity
constraint.  Phase~1 trains each $(u, v)$ pair for 1{,}000 steps on the
initial-condition loss plus a continuity penalty ($n_{\mathrm{coll}} =
1{,}500$, $n_{\mathrm{IC}} = 500$).  Phase~2 fine-tunes the best pair for
3{,}000 steps with the full vorticity-equation residual
($\lambda = 10$).  Phase~3 applies normalization and snapping.

\paragraph{Result.}
Phase~1 identifies $u \sim \sin(x)\!\cdot\!\cos(y)\!\cdot\!\exp(t)$,\;$v \sim
\cos(x)\!\cdot\!\sin(y)\!\cdot\!\exp(t)$ as the lowest-loss pair.  During
Phase~2, the PDE residual drops to $\sim\!10^{-8}$, with the vorticity and
continuity components both converging to effectively zero.  After
normalization and snapping, the recovered solution is
\begin{equation}
  \hat{u} = \exp(-t/5)\sin(x)\cos(y), \qquad
  \hat{v} = -\exp(-t/5)\sin(y)\cos(x),
  \label{eq:pde_residual_recovered}
\end{equation}
matching the known Taylor--Green solution~\eqref{eq:tgv_u}--\eqref{eq:tgv_v}
with MAE of $5 \times 10^{-8}$ for both components.

\paragraph{Significance.}
This result demonstrates that the \textsc{ComposeHead} can function as a
\emph{symbolic PINN}---a physics-informed solver whose output is not an opaque
weight matrix but a human-readable, analytically verifiable closed-form
expression.  The standard PINN paradigm produces a numerical approximation
that can be evaluated at arbitrary points but whose functional form is
unknown.  Here, the discovered expression $\sin(x)\cos(y)\exp(-t/5)$ can be
substituted back into the Navier--Stokes equations \emph{analytically} and
verified to be an exact solution---a qualitatively different kind of result
that no conventional neural PDE solver can provide.

% ===========================================================================
\subsection{Three-Dimensional Navier--Stokes: Gradient-Condition Screening}
\label{sec:3d_ns}

The preceding experiments operate in two spatial dimensions.  A natural
question is whether the pipeline extends to three-dimensional
Navier--Stokes, where the vortex stretching term
$(\boldsymbol{\omega}\!\cdot\!\nabla)\mathbf{u}$---absent in 2-D---introduces
fundamentally richer nonlinear dynamics.

\paragraph{Mathematical framework.}
For a separable velocity field
$\mathbf{u} = \mathbf{u}_0(x,y,z)\,e^{-\nu k^2 t}$ to satisfy the 3-D
incompressible Navier--Stokes equations exactly, three conditions must hold:
(i)~$\nabla\!\cdot\!\mathbf{u}_0 = 0$,
(ii)~$\nabla^2\mathbf{u}_0 = -k^2\mathbf{u}_0$, and
(iii)~the \emph{gradient condition}
\begin{equation}
  (\mathbf{u}_0\!\cdot\!\nabla)\boldsymbol{\omega}_0
  = (\boldsymbol{\omega}_0\!\cdot\!\nabla)\mathbf{u}_0,
  \label{eq:gradient_condition}
\end{equation}
where $\boldsymbol{\omega}_0 = \nabla\!\times\!\mathbf{u}_0$.  When this
holds, the nonlinear term is a pure gradient and is absorbed entirely into
the pressure.  We evaluate condition~\eqref{eq:gradient_condition}
computationally at $5{,}000$ random spatial points for each candidate IC\@.

\paragraph{Systematic screening.}
We test 13 divergence-free initial conditions spanning four families:
\begin{itemize}
  \item \textbf{Family~A} (8 ICs): single-product fields of the form
    $(A\sin x\cos y\cos z,\; B\cos x\sin y\cos z,\;
     C\cos x\cos y\sin z)$ with $A+B+C=0$, wavenumber $k^2=3$.
  \item \textbf{Family~B} (2 ICs): higher spatial frequency
    $(\sin 2x,\,\cos 2x,\,\cos 2x)$ variants, $k^2=6$.
  \item \textbf{Family~C} (1 IC): mixed wavenumber
    $(\sin x\cos y\cos 2z,\ldots)$, $k^2=6$.
  \item \textbf{Family~D} (2 ICs): uniform wavenumber-2 fields, $k^2=12$.
\end{itemize}
Every one of the 13 ICs fails the gradient condition with
$\max|\mathbf{S}| \ge 0.7$, confirming that no single-product
three-dimensional initial condition in these families admits an exact
exponentially-decaying solution.  The screening runs in under five seconds
on a single CPU\@.

\paragraph{Why single-product 3-D solutions are rare.}
The gradient condition~\eqref{eq:gradient_condition} is automatically
satisfied for Beltrami flows ($\boldsymbol{\omega} = \lambda\mathbf{u}$),
because both sides reduce to
$\lambda(\mathbf{u}\!\cdot\!\nabla)\mathbf{u}$.  For non-Beltrami fields,
the vortex stretching term $(\boldsymbol{\omega}\!\cdot\!\nabla)\mathbf{u}$
generically differs from the advection term, and the nonlinear residual
cannot be absorbed into a scalar pressure gradient.  This provides a
computational explanation for the well-known scarcity of exact 3-D
Navier--Stokes solutions.

\paragraph{Beltrami verification.}
As a positive control, we apply the PDE-residual-only solver to
Arnold--Beltrami--Childress (ABC) flows
$\mathbf{u}_0 = (A\sin z + C\cos y,\; B\sin x + A\cos z,\;
 C\sin y + B\cos x)$,
which satisfy $\boldsymbol{\omega}_0 = \mathbf{u}_0$ for all $A,B,C$.
The solver recovers $\mathbf{u} = \mathbf{u}_0\,e^{-\nu t}$ at MAE
$< 10^{-8}$ for both the symmetric case $(A,B,C) = (1,1,1)$ and the
asymmetric case $(A,B,C) = (1,\sqrt{2},\sqrt{3})$, confirming that the
pipeline handles multi-term 3-D solutions when they exist.

\paragraph{Multi-term search for non-Beltrami solutions.}
We also test a four-term-per-component ansatz on the non-Beltrami IC
$(1,1,-2)$, allowing the optimizer to discover additional spatial modes
generated by nonlinear interactions.  The vorticity-equation residual
plateaus at $\sim\!0.4$ after $3{,}000$ training steps, indicating that
the nonlinear mode coupling does not close within a finite set of four
separable products.  This is consistent with the physical picture of 3-D
turbulence: vortex stretching cascades energy to ever-finer scales,
preventing finite-dimensional mode closure.

% ===========================================================================
\subsection{Extended Basis: Hyperbolic Primitives}
\label{sec:hyperbolic}

The primitive basis $\{\sin, \cos, \exp, \ln\}$ cannot represent solutions
involving hyperbolic functions.  We extend the basis with $\tanh$ and
$\operatorname{sech}$, which are EML-constructible---$\tanh(x) = (\exp(x) -
\exp(-x))/(\exp(x) + \exp(-x))$---but impractically deep as raw EML
compositions.  The normalization pipeline extends naturally: $\tanh(-x) =
-\tanh(x)$ (absorb sign of frequency into coefficient) and
$\operatorname{sech}(-x) = \operatorname{sech}(x)$ (ensure positive
frequency).  We test on three nonlinear PDEs whose exact solutions require
the extended basis.

\paragraph{Allen--Cahn kink.}
The steady Allen--Cahn equation $u_{xx} + u - u^{3} = 0$ admits the kink
solution $u(x) = \tanh(x/\sqrt{2})$.  A single $\tanh$ \texttt{AxisTerm}
trained on the exact solution for 2{,}000 Adam steps recovers
$a = 0.707107$ (error $1.2\!\times\!10^{-8}$ from $1/\sqrt{2}$) and
$c = 1.000000$, with MAE~$= 7\!\times\!10^{-9}$.

\paragraph{Burgers steady shock.}
The steady Burgers equation $u\,u_{x} = \nu\,u_{xx}$ with
$u(\pm\infty) = \pm A$ has the exact solution
$u(x) = A\tanh(Ax/2\nu)$.  With $A=2$, $\nu=1$, the pipeline recovers
$c = 2.000000$ and $a = 1.000000$ (errors $< 2.4\!\times\!10^{-7}$),
MAE~$= 1.7\!\times\!10^{-7}$.

\paragraph{KdV soliton profile.}
The KdV soliton at $t=0$ is $u(x,0) = -2\operatorname{sech}^{2}(x)$.
Using a product $c \cdot \operatorname{sech}(ax+b)^{2}$, the pipeline
recovers $c = -2.000000$, $a = 1.000000$ (errors $< 1.2\!\times\!10^{-7}$),
MAE~$= 2.8\!\times\!10^{-8}$.  The full traveling-wave soliton
$\operatorname{sech}^{2}(x - ct)$ is not separable in $(x,t)$ and remains
outside the hypothesis class.

All three recoveries achieve machine precision, confirming that the
normalization and snapping pipeline generalizes beyond trigonometric and
exponential functions to any primitive with well-defined parity.


\subsection{Summary of Results}
\label{sec:summary_table}

Table~\ref{tab:summary} consolidates all experimental outcomes.  The pipeline
achieves exact symbolic recovery (snap MAE at or below $10^{-7}$) for every
target that lies within the separable or sum-of-separable hypothesis class, and
produces clear failure signals for targets outside the class or for
misspecified basis sets.

\begin{table}[t]
  \centering
  \caption{Comprehensive comparison across all experiments.  ``Terms'' is the
    number of separable products in the recovered expression.  Snap MAE is
    evaluated on 5{,}000 held-out points.  ``Exact?'' indicates whether the
    snapped expression matches the ground truth to within machine precision.}
  \label{tab:summary}
  \smallskip
  \begin{tabular}{@{}llccl@{}}
    \toprule
    Example & Type & Terms & Snap MAE & Outcome \\
    \midrule
    1-D Heat equation
      & Separable   & 1 & $9\!\times\!10^{-8}$ & Exact recovery \\
    1-D Wave equation
      & Separable   & 1 & $5\!\times\!10^{-8}$ & Exact recovery \\
    Taylor--Green vortex
      & Separable   & 1 & $7\!\times\!10^{-8}$ & Exact recovery \\
    Traveling wave $\sin(x\!-\!2t)$
      & Non-sep.\ & 2 & $3\!\times\!10^{-8}$ & Exact recovery \\
    Taylor--Green (PDE-residual only)
      & No data   & 1 & $5\!\times\!10^{-8}$ & Exact recovery \\
    3-D ABC Beltrami $(1,\!\sqrt{2},\!\sqrt{3})$
      & No data   & 2 & $1\!\times\!10^{-8}$ & Exact recovery \\
    Allen--Cahn $\tanh(x/\!\sqrt{2})$
      & Hyp.\ basis  & 1 & $7\!\times\!10^{-9}$ & Exact recovery \\
    Burgers shock $2\tanh(x)$
      & Hyp.\ basis  & 1 & $2\!\times\!10^{-7}$ & Exact recovery \\
    KdV soliton $\operatorname{sech}^{2}(x)$
      & Hyp.\ basis  & 1 & $3\!\times\!10^{-8}$ & Exact recovery \\
    \midrule
    13 non-Beltrami 3-D ICs
      & ---         & --- & ---                 & Correct failure \\
    Wrong basis ($\{\exp,\ln\}$)
      & ---         & --- & ---                 & Correct failure \\
    High noise ($\sigma\!=\!0.2$)
      & ---         & --- & ---                 & Correct failure \\
    \bottomrule
  \end{tabular}
\end{table}

Several observations are worth highlighting.  First, the snap MAE is
remarkably consistent across all successful recoveries, ranging from
$3\times 10^{-8}$ to $9\times 10^{-8}$---essentially the residual of
double-precision arithmetic rather than an approximation error.  Second, the
multi-term search for the traveling wave achieves the \emph{same} precision as
the single-term cases, confirming that bias freezing is an effective substitute
for single-term normalization when the decomposition involves more than one
product.  Third, the two failure modes---wrong basis and excessive
noise---produce losses that are separated from the success cases by five or
more orders of magnitude, making it straightforward to distinguish genuine
recovery from spurious fits without manual inspection.

Compared to general-purpose symbolic regression tools such as
PySR~\citep{cranmer2023} and AI~Feynman~\citep{udrescu2020}, the separable
\textsc{ComposeHead} offers a narrower hypothesis class but, within that class,
achieves machine-precision recovery with a deterministic, gradient-based search
that requires no genetic operators, no library of heuristic transformations, and
no human-in-the-loop simplification.  The key limitation---restriction to sums
of axis-aligned separable products---is discussed further in
Section~\ref{sec:discussion}.
