\section{Introduction}
\label{sec:introduction}

Partial differential equations govern phenomena across fluid dynamics, quantum
mechanics, and continuum mechanics, yet obtaining analytical solutions remains
one of the most difficult tasks in applied mathematics.  When closed-form
solutions do exist, they provide qualitative insight, asymptotic guarantees,
and computational efficiency that no numerical approximation can match.
The recent success of deep-learning-based PDE solvers has dramatically
expanded the class of problems that can be tackled computationally, but it has
done so at the cost of interpretability: a trained physics-informed neural
network (PINN)~\cite{raissi2019} produces a set of floating-point weights from
which no human-readable formula can be extracted.

\paragraph{Neural PDE solvers: accuracy without insight.}
PINNs and their descendants embed the governing PDE into the loss function of a
neural network and optimize over the network parameters so that the learned
map approximately satisfies the differential equation together with its
boundary and initial conditions.  This paradigm is flexible---it requires
neither a mesh nor a specialized numerical scheme---and has yielded
impressive results on forward and inverse problems alike.  However, the output
of such a solver is an opaque parameterization: one obtains a weight matrix,
not a formula.  Post-hoc interpretability techniques can highlight which input
regions matter, but they cannot recover the symbolic structure that a
mathematician would recognize as a solution.

\paragraph{Symbolic regression: powerful but limited in scope.}
An alternative line of work seeks to discover symbolic expressions directly from
data.  Tools such as PySR~\cite{cranmer2023} and AI~Feynman~\cite{udrescu2020}
have achieved striking success on algebraic physical laws---rediscovering
conservation laws, constitutive relations, and dimensionless
groupings from noisy measurements.  These methods search over a grammar of
elementary operations and fit scalar expressions of moderate complexity.
Yet PDE solutions pose a qualitatively different challenge.  A typical
analytical solution is not a single algebraic expression in one variable but a
\emph{product of functions of different variables}---for instance,
$\sin(x)\cos(y)\exp(-\nu t)$---and the combinatorial explosion of
multi-variable compositions quickly overwhelms generic symbolic regression
engines.

\paragraph{Elementary functions as a principled basis.}
A natural question arises: which set of primitive functions should a symbolic
search use?  An ad hoc selection risks omitting functions needed for a given
solution class.  The recent universality result of
Odrzywo{\l}ek~\cite{odrzywołek2026} provides a principled answer.
Odrzywo{\l}ek proved that the two-argument function
$\mathrm{eml}(x,y) = e^{x} - \ln y$ is \emph{universal}: every computable
function can be represented as a finite composition of $\mathrm{eml}$ with
itself.  Although the raw EML representation is unwieldy---multiplication alone
requires a reverse-Polish representation of length~41---the result has a
practically important corollary: the elementary functions $\{\sin, \cos, \exp,
\ln\}$ can each be constructed from EML compositions via algebraic identities.
This establishes the set $\{\sin, \cos, \exp, \ln\}$ not as an arbitrary
library but as a \emph{verified basis} derived from a universal primitive.
Moreover, this function class is closed under partial differentiation, a
property that is essential when the loss function involves PDE residuals
computed via automatic differentiation.

\paragraph{Our approach: normalization-driven symbolic recovery.}
The central challenge in recovering exact symbolic expressions from
gradient-trained models is not the choice of architecture but rather the
\emph{extraction} of clean symbolic forms from continuously parameterized
representations.  Elementary functions exhibit gauge freedoms that confound
naive coefficient readout: a trained model may represent $\cos(x)$ as
$\sin(x + \pi/2)$, absorb constant factors into exponential biases, or split a
single physical term across multiple basis functions with nearly cancelling
coefficients.  We address these difficulties with two key ideas.

First, we introduce a \emph{normalization pipeline} that resolves the gauge
freedom inherent in parameterized elementary functions.  The pipeline absorbs
exponential biases into linear coefficients, canonicalizes trigonometric phases
by exploiting the identities $\sin(x + \pi/2) = \cos(x)$ and
$\cos(x + \pi) = -\cos(x)$, and rounds the resulting coefficients to nearby
rational or otherwise recognizable values.  This normalization step is the
primary enabler of exact symbolic recovery, improving coefficient accuracy by
approximately five orders of magnitude relative to direct readout of trained
parameters.

Second, we adopt a \emph{single-term search strategy} in which the model is
trained with $K=1$ active term at a time, avoiding the signal-splitting
pathology in which a multi-term ansatz distributes a single physical mode across
several basis functions with partially cancelling coefficients.  Multi-term
solutions are assembled by sequential residual fitting.

These ideas are realized in a differentiable architecture we call the
\textsc{ComposeHead}, which represents candidate solutions as trainable linear
combinations of \emph{separable terms}:
\begin{equation}
  \label{eq:composehead-intro}
  \hat{u}(\mathbf{x}) \;=\; \sum_{k=1}^{K} c_k \prod_{d=1}^{D}
    f_{k,d}\!\bigl(\alpha_{k,d}\, x_d + \beta_{k,d}\bigr),
\end{equation}
with each factor $f_{k,d}$ drawn from the primitive library
$\{\sin,\cos,\exp,\ln,\mathrm{id}\}$ and each $\alpha_{k,d}$, $\beta_{k,d}$,
$c_k$ a trainable scalar coefficient.  The separable product structure mirrors
the classical method of separation of variables, which has been a cornerstone of
analytical PDE theory for three centuries.  By restricting the hypothesis class
to this well-understood form, the optimization problem reduces from learning
arbitrary neural-network weights to identifying which elementary functions
compose along each coordinate axis and with what coefficients.

\paragraph{Scope and limitations.}
The method proposed here is best understood as \emph{symbolic regression
specialized to separable PDE solutions}.  The ansatz in
Equation~\eqref{eq:composehead-intro} is expressive for the broad class of
problems whose analytical solutions factor across coordinate dimensions---a
class that includes many canonical problems in mathematical physics---but it
does not address existence, uniqueness, or regularity questions, nor does it
replace mesh-based solvers for problems lacking closed-form solutions.  Problems
whose solutions are inherently non-separable (e.g., those requiring similarity
variables or integral representations) fall outside the current scope.  Our
claims are accordingly conservative: we demonstrate that, when an analytical
solution exists and has separable structure, the method can recover it exactly
from PDE-residual supervision alone.

\paragraph{Contributions.}
The principal contributions of this work are as follows.
\begin{itemize}
  \item \textbf{Normalization and snapping pipeline.}  We introduce a
    post-training procedure that resolves the gauge freedom in parameterized
    elementary functions---absorbing exponential biases, canonicalizing
    trigonometric phases, and snapping coefficients to exact values---converting
    a continuously parameterized model into a closed-form symbolic expression.
    This pipeline yields approximately five orders of magnitude improvement in
    coefficient accuracy relative to direct parameter readout.

  \item \textbf{Single-term search strategy.}  We show that training with one
    active term at a time, followed by sequential residual fitting, avoids the
    signal-splitting pathology that arises in multi-term optimization and is
    essential for reliable symbolic extraction.

  \item \textbf{ComposeHead architecture with separable ansatz.}  We introduce a
    differentiable architecture whose hypothesis class is the set of linear
    combinations of separable products of axis-aligned elementary functions,
    enabling gradient-based search over symbolic PDE solution candidates.

  \item \textbf{EML-grounded primitive basis.}  We provide differentiable,
    algebraically verified implementations of $\exp$, $\ln$, $\sin$, and $\cos$
    built from compositions of the EML function of~\cite{odrzywołek2026},
    establishing a principled (rather than ad hoc) justification for the
    primitive library and ensuring differentiation closure of the function class.

  \item \textbf{Exact recovery across PDE classes.}  We demonstrate the
    pipeline on four PDE benchmarks---1-D heat, 1-D wave, the non-separable
    traveling wave $\sin(x-2t)$, and the 2-D incompressible Navier--Stokes
    equations---recovering every target at machine precision (MAE ${<}\,10^{-7}$).

  \item \textbf{PDE-residual-only symbolic discovery.}  We show that the
    pipeline can discover exact analytical solutions from the governing
    equations and initial conditions alone, with no solution data whatsoever.
    Applied to the 2-D Navier--Stokes equations, it recovers the Taylor--Green
    vortex $u = \sin(x)\cos(y)\exp(-t/5)$,\;$v = -\cos(x)\sin(y)\exp(-t/5)$
    at MAE $5\times 10^{-8}$---functioning as a symbolic PINN whose output is
    a verifiable closed-form expression rather than an opaque weight matrix.
\end{itemize}

\noindent The remainder of the paper is organized as follows.
Section~\ref{sec:background} reviews the EML universality result and the
relevant symbolic regression literature.
Section~\ref{sec:method} details the primitive constructions, the
\textsc{ComposeHead} architecture, and the normalization pipeline.
Section~\ref{sec:experiments} presents the Taylor--Green vortex experiment and
ablation studies.
Section~\ref{sec:discussion} discusses limitations and directions for future
work.
